### 2012-10-03 2.tex
Edits to 2 and 4, and writing in 3

\citet{Halle1962} and \citet{Chomsky1965} observe that speakers internalize generalizations about possible and impossible words in their language. \citeauthor{Chomsky1965} illustrate this knowledge with two nonce words \emph{blick} [blɪk] and \emph{bnick} [bnɪk]. Whereas \emph{blick} is unattested, but in some sense a ``possible'' word of English, speakers immediately recognize \emph{bnick} to be ill-formed, an impossible word of the language.

A naïve account of this contrast is based on the assumption that segments or timing units must either be parsed into prosodic structures like the syllable, or be subject to phonological repair. Early precedents for this can be found in dissertations by \citet[10f.]{Hooper1973} and \citet[57f.]{Kahn1976}, and it is a fundamental component of many later frameworks \citep[e.g.,][]{Ito1989a,Noske1992,OT}.

@@

Many linguists who have concerned themselves with ``wordlikeness'' now regard the above as overly simplistic, for, as \citeauthor{Shademan2006} argues, such accounts draw a categorical boundary between the possible and impossible which fails to match the granularity at which speakers may report ``wordlikeness'' intuitions.

\begin{quote}
A defect of current grammatical acounts of phonotactics is that they render simple up-or-down decisions concerning well-formedness and cannot account for gradient judgements. But when judgements are elicited in a controlled fashion from speakers, they always emerge as gradient, including all intermediate values. \citep[][371]{Shademan2006} 
\end{quote}

This chapter focuses on the hypothesis, implicit in \citeauthor{Shademan2006}'s statement, that there is an inherent gradient or continuum between possible and impossible words, against an null hypothesis that the observed patterns of judgements are little more than an artifact of the wordlikeness task itself. \S\ref{2background} places this debate in a historical perspective, and argues that the naïve but widely accepted form of this hypothesis fails the test of falsifiability. In \S\ref{2evaluation}, a quantitative evaluation of a revised version of this hypothesis reveals a poor match between the gradience in speakers' judgements and contemporary computational models of wordlikeness. Taken together, the findings argue against the gradience hypothesis.

@@

However, not all early literature is concerned with gradience. \citet[][31]{Vogt1954}, for instance, recognizes that the taxonomic phoneme is insufficient to account for many wordlikeness contrasts. \citeauthor{Vogt1954} observes that allophony may account for the absence of certain phone sequences, but it does not provide a suitable explanation for the absence of initial [bn] in English, nor does it make correct predictions about the surface realization of an underlying initial /bn/. \citeauthor{Vogt1954} concludes that additional grammatical machinery will be needed to account for possible and impossible words.

@@

Gradient grammaticality is assumed in the earliest work on generative syntax, \emph{The logical structure of linguistic theory}, in which \citet[][132]{LSLT} writes that ``there is little doubt that speakers can fairly consistently order new utterances, never previously heard, with respect to their degree of `belongingness' to the language''. Later, \citet{ASPECTS} proposes that different syntactic violations give rise to different degrees of ungrammaticality. Unlike \citeauthor{Fischer-Jorgensen1952}, however, for \citeauthor{LSLT} this continuum does incorporate a hard-and-fast boundary between the grammatical and ungrammatical: there are no constrasts between grammatical utterances, merely between different degrees of ungrammaticality. For \citeauthor{LSLT}, all grammatical utterances are alike with respect to their grammaticality, but there are many degrees of ungrammaticality \citep[][61]{Schutze1996}.

\citet{Halle1962} and \citet{Stanley1967} concern themselves only with categorical wordlikeness contrasts, however. Both authors derive wordlikeness effects from morphophonemic redundancy rules acting on underlying representations. The discussion of \emph{bnick} by \citet[][101]{Chomsky1965} is a good example of this type of analysis. \citeauthor{Chomsky1965} observe that a consonant before a word-initial stop can only be a liquid in English. A corresponding redundancy rule which specifies a consonant in this position as [$+$\textsc{Liquid}] precludes the derivation of \emph{bnick}, since there is no way to realize the second consonant as a nasal.\footnote{This is not the only redundancy that could rule out \emph{bnick}. For instance, /s/ is the only word-initial obstruent that may precede a nasal consonant in English.}

@@

In \emph{The sound pattern of English} (Henceforth \emph{SPE}), \citet{SPE} argue that a complete solution to wordlikeness must countenance gradience. In addition to the contrast of \emph{blick} and \emph{bnick}, they introduce a third nonce word \emph{bnzk}, which they take to be even less well-formed than ``impossible'' \emph{bnick}.

@@

In some languages (e.g., Imdlawn Tashlhiyt Berber [tm.zħ] `she jested', [tzmt] `it (f.) is stifling'; \citealt{Dell1985}), \emph{bnzk} is syllabifiable, so its ill-formedness is once again a property of English, not a universal.

@@

Despite the considerable attention given to the proposals of \emph{SPE} in the wake of that book's publication in 1968, the \emph{SPE} wordlikeness model has received almost no attention in the literature. At the risk of explaining what might be no more than an an accidental gap in the historical record, the novel aspects of the \emph{SPE} model---gradience derived from similarity to existing lexical entries---seem to have been overshadowed by compelling arguments against \citeauthor{SPE}'s assumption \citep[see also][]{Halle1962,Stanley1967} that wordlikeness contrasts derive from properties of underlying forms. \citet{Shibatani1973} observes that there are some generalizations about surface forms which give rise to wordlikeness contrasts, but cannot be stated as constraints on underlying forms. An example from German is shown in (\ref{fd}) below.

@@

\noindent The plural appears to preserve a contrast in final obstruent voicing which is absent in the singular:\footnote{While there is a contentious debate as to whether devoicing in German is completely neutralizing \citep[e.g.,][]{Fourakis1984}) or not \citep[e.g.,][]{Port1985}, it is irrelevant to the discussion at hand.} word-final voiced stops are never found in German, and \citet[95]{Shibatani1973} reports that ``it is easy to show that a native speaker of German rejects those forms \emph{on the ground that they end in voiced obstruents}'' (emphasis in original). However, given that root-final voicing is not predictable (e.g., /ɡʀaːt \alt{} ɡʀaːd/), the process of \textsc{Final Devoicing} cannot be any kind of lexical redundancy and is mysterious under the \emph{SPE} account. The alternative proposed by \citeauthor{Shibatani1973} and by \citet{Clayton1976}, is that to say that nonce words like [ɡʀaːd] are ill-formed not beacuse of any underlying property, but because they fail to undergo an otherwise-exceptionless phonological process, \textsc{Final Devoicing}, or equivalently, the surface-true generalization it derives.

@@

The arguments of \citeauthor{Shibatani1973} and others led theorists to focus their attention on properties of surface representations as determinants of wordlikeness. Though syllabification plays no role in \emph{SPE}, it is crucial to many earlier studies \citep[see][]{Goldsmith2011b}, and it received particular attention in the 1970s. \citet{Hooper1973} and \citet{Kahn1976} argue that the syllable is useful for defining wordlikeness generalizations.\footnote{\citet{Steriade1999} and \citet{Blevins2003}, however, argue that a number of phonotactic generalizations previously stated in syllabic terms can be reanalyzed without making reference to syllables.} \citeauthor{Hooper1973} argues, for instance, that [bn], impossible as an English onset, is unobjectionable as a syllable contact cluster in nonce words like \emph{stabnik} (or in names like \emph{Abner}), and that this demonstrates the superiority of syllable-based wordlikeness generalizations. This already signals further trouble for alternative accounts which focus on underlying forms. Syllabification may span morphs, is generally predictable, and is universally non-contrastive, and as a consequence, few posit in to be present in underlying representations \citep[though see, e.g.,][]{Vaux2003}. \citeauthor{Hooper1973} also points to loanword adaptations which produce native syllable structure \citep[e.g.,][]{Carlisle1991} as evidence that syllabification is part of the phonological computation. Further enrichments to the theory are provided by the autosegmental theory of the syllable \citep{McCarthy1979b}, which envisions the syllable as an articulated tree structure \citep[as envisioned by][]{Pike1947a}, and theories like prosodic licensing \citep{Ito1989a}, in which syllabification triggers phonological repairs.

@@

Most relevant to the question at hand, \citet{Vitevitch1997} and \citet{Frisch2000} find that speakers' wordlikeness ratings of multisyllabic words are correlated wtih the positional probailities of the constituent syllables. Unfortunately, none of these researchers make any effort to eliminate the possibility that the low positional probability stimuli are ``impossible'' words of English. In fact, 
Chapter \ref{clusters} argues
%the author has argued elsewhere \citep{Gorman2012c}
that many of the stimuli used by \citet{Vitevitch1997} and \citet{Frisch2000} contain illicit word-medial consonant clusters. While \citeauthor{Vitevitch1997} neither control nor manipulate the well-formedness of medial clusters, in a post-hoc test they consider a probabilistic measure of cluster well-formedness, which reveals that cluster well-formedness is correlated with syllable-internal positional probabilities and the wordlikeness judgements. \citeauthor{Vitevitch1997} ultimately conclude that it cannot explain all the variation in wordlikeness.

The aforementioned studies all conclude that the gradient measure of positional probability correlates with behavioral results. As the flaws of the \citet{Vitevitch1997} study demonstrate, the aforementioned studies do little to tease apart the gradient and categorical aspects of phonotactics. More generally, they do little to distinguish between positional probability and closely correlated measures like bigram probability (see \S\ref{bigram} below) or neighborhood density, since these studies carefully select stimuli which either have high or low values for all of positional probability, bigram probablility, and neighborhood density. This is particularly troublesome given that positional probability was created \emph{ex nihilo}; in contrast, the effects of neighborhood density in various psycholinguistic tasks are emergent properties of many models of speech production \citep[e.g.,][]{Luce1998,Luce2000} and perception \citep{Marslen-Wilson1984,Marslen-Wilson1987,McClelland1986,Norris1994,Norris2000}.

@@

This proposition has the form of an indicative conditional. While the antecedent of this condition, the presence or absence of intermediate ratings in a graded judgement task, is readily observable, it has no bearing on the truth or falsity of the conditional proposition. However, whereas a false consequent (e.g., an all-or-nothing concept) would potentially falsify the proposition, the consequent, a mental state, cannot be directly observed. If this is the case, then both the proposition and the consequent (the proposition that wordlikeness is itself gradient) fail the test of falsification, placing it beyond the reach of scientific inquiry.

@@

\citeauthor{Armstrong1983} note that if subjects produce intermediate ratings for these categories, then the naïve proposition is false, and this is what \citeauthor{Armstrong1983} find in their experiments. They ask subjects to rate, on a seven-point scale, the extent to which, e.g., certain odd counting numbers represent the concept ``odd number'', and find that speakers do make use of intermediate values, as can be seen in Figure \ref{agg}. In summary, the proposition that intermediate ratings indicate mental gradience is either untestable, or it is false, and is thus rejected.

@@

\citeauthor{Armstrong1983} do not view their results as evidence that subjects have an internal model of odd numbers which is at odds with the formal, all-or-nothing definition, and in fact, several additional experiments show that subjects deploy the extension of the formal definition in other tasks. Reviewing this study, \citet[215]{Schutze2011} writes that the experiments show that judgements are ``sensitive to factors other than our underlying competence''. In the case of wordlikeness, a case can be made that one such factor is similarity to existing words (e.g., \citealt[][151, fn. 27]{LSLT}, \citealt{Greenberg1964}, \citealt{Vitz1973}, \citealt{Ohala1986b}, \citealt{Sendlmeier1987}, \citealt{Bailey2001}, \citealt{Schutze2005}). However, the mere fact that subjects use intermediate ratings does not show that they do so with any systematicity, or that, e.g., the contrast between \emph{447} and \emph{7} with respect could be given any satisfying explanation. \citeauthor{Schutze2011} further suggests that gradience responses is a property of tasks, not concepts.

\begin{quote}
Putting it another way, when asked for gradient responses, participants will find some way to oblige the experimenter; if doing so is incompatible with the experimenter's actual question, they apparently infer that she must have really intended to ask something slightly different. \citep[][215]{Schutze2011}
\end{quote}

This introduces a troubling possibility, that subjects ``oblige'' experimenters by introducing random noise to their categorical judgements when presented with a Likert task. Were this the case, the observed intermediate judgements would scarcely be worthy of modeling. The alternative considered here is that there exists some model in which the gradience in ratings is systematic.

@@

\noindent There are now many computational models of wordlikeness which proport to do just this. An unfortunate defect of prior evaluations of these models, however, is that they adopt the naïve gradience hypothesis. \citet[382]{Hayes2008a} are just some of the authors who consider the ability to model gradient intuitions to be so important that they do not include any categorical models in their evaluation. As a result, the literature contains no serious attempt to evaluate the gradience hypothesis.

@@

\citet{Greenberg1964} investigated wordlikeness using the technique of free magnitude estimation, a mechanism which has become increasingly popular among syntacticians \citep[e.g.,][]{Bard1996}. At the beginning of the experiment, the subject heard a recording of the word \emph{stick}. In subsequent trials, the subjects heard a nonce word and were asked to report ``how far would you say that is from English?'', with \emph{stick} at ``1''; subjects are told that a word that is ``twice as far from English'' as \emph{stick} should be scored ``2''. The data used here are from \citeauthor{Greenberg1964}'s Experiment B, in which 17 undergraduates were presented 17 stimuli in all. In addition to \emph{stick}, the stimuli include three other English words; these four items were excluded from further analyses, leaving 13 stimuli. As is standard practice in psychophysics \citep[e.g.,][]{Butler1987}, these magnitude ratings were log-transformed before analysis.

@@

@@

The binary baseline used here is a crude implementation of the null hypothesis that there are no gradient effects in wordlikeness judgements. To create such a baseline, it is necessary to distinguish between nonce words which contain a gross phonotactic violation and those which do not. As all stimuli here are monosyllables, this task can be further simplified by separately considering the two major subcomponents of the syllable, the onset and rime. Speakers are particularly adept as separating onsets from rimes \citep{Treiman1986,Treiman1995,Fowler1993}, and a large portion of phonotactic generalizations and violations can be localized to either onset or rime \citep[e.g.,][]{Fudge1969,Treiman1995,Kessler1997,Treiman2000}.

The baseline considers a nonce syllable to be well-formed if it consists of both an attested onset and an attested rime; the free combination of these two components is the only mechanism by which this model can generalize beyond attested words. It is surely the case that this is an ultimately insufficient model: \citet{Albright2009a} observes, for instance, that [ɛsp] is a well-formed rime, even though it is found in no word of English. This problem is more acute in languages with more permissive syllable structures than those of English, and thus a sparser lexicon with regard to phonotactic constraints. For instance, \citet{Fischer-Jorgensen1952} and \citet{Vogt1954} assert that there are many accidental gaps (i.e., possible but unattested structures) in the inventory of consonant clusters in Georgian, a language which admits as many as five adjacent consonants.\footnote{
%Chapter \ref{clusters} 
\citet{Gorman2012c}
reports a similar result for English syllable contact clusters.} Despite these defects, the baseline outperforms the gradient models in most contexts.

\begin{figure} \centering
\includegraphics{density.pdf}
\caption{Average ratings of individual nonce words, linearly transformed to the interval [0, 1], tend to clump into two groups with little overlap; words which consist of  attested onsets and rimes receive ratings near ceiling, whereas ratings of phonotactically invalid words are spread across the lower half of the spectrum.}
\label{dsn}
\end{figure}

@@

\citeauthor{Hayes2008a} (\citeyear{Hayes2008a}; henceforth H\&W) develop a sophisticated model of phonotactic grammaticality which estimates a probability distribution over phoneme sequences by weighing constraints according to the principle of maximum entropy, following \citet{Goldwater2003} and \citet{Jager2007}. H\&W find that the predictions of their model are closely correlated with the \citet{Scholes1966} wordlikeness ratings. A direct replication of their predictions was attempted by using the software, model parameters, and training data as described in that study. Since the training of the maximum entropy model is inherently stochastic, producing slightly different outcomes on each run, the lowest scoring of ten runs is reported (H\&W:396), though in general there is not a great deal of variation between individual runs. One limitation of this model is that it is not feasible to score whole words, as the number of constraints which must be inspected grows exponentially as the size of possible constraints increases. Following H\&W and of \citet{Albright2009a}, who also applies the maximum entropy model to the \citet{Albright2003b} norms, the model is trained and scored only on stimulus onsets. However, as a consequence, the maximum entropy model performs particularly poorly on this data set, as many stimuli contain phonotactic violations in other positions.

@@

\noindent Bigram models are widely used in natural language processing, and \citet{Albright2009a} considers their relevance to modeling wordlikeness judgements. While the focus of \citeauthor{Albright2009a}'s study is on developing a model which uses bigrams over phonological features rather than segments themselves, \citeauthor{Albright2009a}'s evaluation, which includes both the \citeauthor{Scholes1966} and \citeauthor{Albright2003b} data sets, finds an advantage for segmental bigrams.

\citeauthor{Albright2009a} estimates bigram probabilities using the method of maximum likelihood over types in the lexicon. The variant of segmental bigrams used here computes probabilities with a simple type of smoothing in which the count of all possible bigrams (including those never observed) are incremented by one. This technique is known as Laplace, or ``add one'' smoothing. This has the desirable effect that no nonce word is ever assigned a zero probability, and produces in a small increase in the correlation between the \citeauthor{Albright2003b} wordlikeness norms compared with the maximum likelihood estimate (Table \ref{albrightimproved}). For all three data sets, this model also consistenly outperforms positional probability models as defined by \citet{Vitevitch2004} and \citet{Vaden2009}. Given that these probabilities are highly correlatd with bigram probabilities \citep[][54]{Vitevitch1997}, they are not considered further. The bigram model consistently performs well in all the evaluations, and has the highest Spearman correlation with the \citeauthor{Greenberg1964} and \citeauthor{Scholes1966}, and is frequently second place model to the binary baseline elsewhere.

>>>>

\citet{Halle1962} and \citet{Chomsky1965} use two nonce words, \emph{blick} [blɪk] and \emph{bnick} [bnɪk], to illustrate that speakers internalize generalizations about possible and impossible words in their language. While \emph{blick} is unattested, but in some sense a ``possible'' word of English, \citeauthor{Chomsky1965} observe that speakers immediately recognize \emph{bnick} to be ill-formed, an impossible word, despite the fact that it is well-formed in other languages (e.g., Moroccan Arabic \emph{bniqa} `closet').

A naïve account of this contrast is based on the assumption that segments or timing units must either be parsed into prosodic structures like the syllable, or be subject to phonological repair. Early precedents for this assumption can be found in work by \citet[10f.]{Hooper1973} and \citet[57f.]{Kahn1976}, and it is fundamental to many later frameworks \citep[e.g.,][]{Ito1989a,Noske1992,OT}.

@@

However, many linguists now regard the above as overly simplistic, for, as \citeauthor{Shademan2006} argues, such accounts draw a categorical boundary between the possible and impossible which fails to match the granularity at which speakers may report ``wordlikeness'' intuitions.

\begin{quote}
A defect of current grammatical acounts of phonotactics is that they render simple up-or-down decisions concerning well-formedness and cannot account for gradient judgements. But when judgements are elicited in a controlled fashion from speakers, they always emerge as gradient, including all intermediate values. \citep[371]{Shademan2006} 
\end{quote}

This chapter focuses on the hypothesis, implicit in \citeauthor{Shademan2006}'s statement, that there is an inherent gradient or continuum between possible and impossible words, against an null hypothesis that the observed patterns of judgements are little more than an artifact of the wordlikeness task itself. \S\ref{2background} places this debate in a historical perspective, and argues that the naïve but widely accepted form of this hypothesis cannot be falsified. In \S\ref{2evaluation}, a quantitative evaluation of a revised version of this hypothesis reveals a poor match between the gradience in speakers' judgements and contemporary computational models of wordlikeness. Taken together, the findings undermine the gradience hypothesis.

@@

However, not all early literature is concerned with gradience. \citet[31]{Vogt1954}, for instance, recognizes that the taxonomic phoneme is insufficient to account for many wordlikeness contrasts. \citeauthor{Vogt1954} observes that allophony may account for the absence of certain phone sequences, but it does not provide a suitable explanation for the absence of initial [bn] in English, nor does it make correct predictions about the surface realization of an underlying initial /bn/. \citeauthor{Vogt1954} concludes that additional grammatical machinery will be needed to account for possible and impossible words.

@@

Gradient grammaticality is assumed in the earliest work on generative syntax, \emph{The logical structure of linguistic theory}, in which \citet[132]{LSLT} writes that ``there is little doubt that speakers can fairly consistently order new utterances, never previously heard, with respect to their degree of `belongingness' to the language''. Later, \citet{ASPECTS} proposes that different syntactic violations give rise to different degrees of ungrammaticality. Unlike \citeauthor{Fischer-Jorgensen1952}, however, for \citeauthor{LSLT} this continuum does incorporate a hard-and-fast boundary between the grammatical and ungrammatical: there are no constrasts between grammatical utterances, merely between different degrees of ungrammaticality. For \citeauthor{LSLT}, all grammatical utterances are alike with respect to their grammaticality, but there are many degrees of ungrammaticality \citep[61]{Schutze1996}.

\citet{Halle1962} and \citet{Stanley1967} concern themselves only with categorical wordlikeness contrasts, however. Both authors derive wordlikeness effects from morphophonemic redundancy rules acting on underlying representations. The discussion of \emph{bnick} by \citet[101]{Chomsky1965} is a good example: \citeauthor{Chomsky1965} observe that a consonant before a word-initial stop can only be a liquid in English. A corresponding redundancy rule which specifies a consonant in this position as [$+$\textsc{Liquid}] precludes the derivation of \emph{bnick}, since there is no way to realize the second consonant as a nasal.\footnote{This is not the only redundancy that could rule out \emph{bnick}. For instance, /s/ is the only word-initial consonant that may precede a nasal consonant in English.}

@@

In \emph{The sound pattern of English} (Henceforth \emph{SPE}), \citet{SPE} argue that a complete solution to wordlikeness must countenance gradience. In addition to the contrast of \emph{blick} and \emph{bnick}, they introduce a third nonce word \emph{bnzk}, which they take to be even less well-formed than ``impossible'' \emph{bnick}. They write:

@@

%In some languages (e.g., Imdlawn Tashlhiyt Berber [tm.zħ] `she jested', [tzmt] `it (f.) is stifling'; \citealt{Dell1985}), \emph{bnzk} is syllabifiable, so its ill-formedness is once again a property of English, not a universal.

@@

Despite the considerable attention given to the proposals of \emph{SPE} in the wake of that book's publication in 1968, the \emph{SPE} wordlikeness model has received almost no attention in the literature. At the risk of explaining what might be no more than an an accidental gap in the historical record, the novel aspects of the \emph{SPE} model---gradience derived from similarity to existing lexical entries---seem to have been overshadowed by compelling arguments against \citeauthor{SPE}'s assumption \citep[see also][]{Halle1962,Stanley1967} that wordlikeness contrasts derive solely from properties of underlying forms. \citet{Shibatani1973} observes that there are some generalizations about surface forms which give rise to wordlikeness contrasts, but cannot be stated as constraints on underlying forms. An example from German is shown in (\ref{fd}) below.

@@

\noindent The plural appears to preserve a contrast in final obstruent voicing which is absent in the singular:\footnote{While there is a contentious debate as to whether devoicing in German is completely neutralizing \citep[e.g.,][]{Fourakis1984} or not \citep[e.g.,][]{Port1985}, it is irrelevant to the discussion at hand.} word-final voiced stops are never found in German, and \citet[95]{Shibatani1973} reports that ``it is easy to show that a native speaker of German rejects those forms \emph{on the ground that they end in voiced obstruents}'' (emphasis in original). However, given that root-final voicing is not predictable (e.g., /ɡʀaːt \alt{} ɡʀaːd/), the process of \textsc{Final Devoicing} cannot be any kind of lexical redundancy and is mysterious under the \emph{SPE} account. The alternative proposed by \citeauthor{Shibatani1973} and by \citet{Clayton1976}, is that to say that nonce words like [ɡʀaːd] are ill-formed not because of any underlying property, but since they fail to undergo an otherwise-exceptionless phonological process, \textsc{Final Devoicing}, or equivalently, the surface-true generalization it derives.

@@

The arguments of \citeauthor{Shibatani1973} and others led theorists to focus their attention on properties of surface representations as determinants of wordlikeness. Though syllabification plays no role in \emph{SPE}, it is crucial to many earlier studies \citep[for a review, see][]{Goldsmith2011b}, and it received particular attention in the 1970s. \citet{Hooper1973} and \citet{Kahn1976} argue that the syllable is useful for defining wordlikeness generalizations.\footnote{\citet{Steriade1999} and \citet{Blevins2003}, however, argue that a number of phonotactic generalizations previously stated in syllabic terms can be reanalyzed without making reference to syllables.} \citeauthor{Hooper1973} argues, for instance, that [bn], impossible as an English onset, is unobjectionable as a syllable contact cluster in nonce words like \emph{stabnik} (or in names like \emph{Abner}), and that this demonstrates the superiority of syllable-based wordlikeness generalizations. This already signals further trouble for alternative accounts which focus on underlying forms. Syllabification may span morphs, is generally predictable, and is universally non-contrastive, and as a consequence, few posit in to be present in underlying representations \citep[though see, e.g.,][]{Vaux2003}. \citeauthor{Hooper1973} also points to loanword adaptations which produce native syllable structure \citep[e.g.,][]{Carlisle1991} as evidence that syllabification is part of the phonological computation. Further enrichments to the theory are provided by the autosegmental theory of the syllable \citep{McCarthy1979b}, which envisions the syllable as an articulated tree structure \citep[as first envisioned by][]{Pike1947a}, and theories like prosodic licensing \citep{Ito1989a}, in which syllabification triggers phonological repairs.

@@

Most relevant to the question at hand, \citet{Frisch2000} and \citet{Vitevitch1997} find that speakers' wordlikeness ratings of multisyllabic words are correlated wtih the positional probailities of the constituent syllables. Unfortunately, none of these researchers make any effort to eliminate the possibility that the low positional probability stimuli are ``impossible'' words of English. In fact, 
Chapter \ref{clusters} argues
%the author has argued elsewhere \citep{Gorman2012c}
that many of the stimuli used by \citeauthor{Frisch2000} and \citeauthor{Vitevitch1997} contain illicit word-medial consonant clusters. While \citeauthor{Vitevitch1997} neither control nor manipulate the well-formedness of medial clusters, in a post-hoc test they consider a probabilistic measure of cluster well-formedness, which reveals that cluster well-formedness is correlated with syllable-internal positional probabilities and wordlikeness judgements, but \citeauthor{Vitevitch1997} ultimately conclude this cannot explain all the variation in wordlikeness.

The aforementioned studies all conclude that the gradient measure of positional probability correlates with behavioral results. As the flaws of the \citet{Vitevitch1997} study demonstrate, the aforementioned studies do little to tease apart the gradient and categorical aspects of phonotactics. More generally, they do little to distinguish between positional probability and closely correlated measures like bigram probability (see \S\ref{bigram} below) or neighborhood density, since these studies carefully select stimuli which either have high or low values for all of positional probability, bigram probablility, and neighborhood density. This is particularly troublesome given that no justification has ever been given for the positional probability measure in the first place; it appears to have been created \emph{ex nihilo}; in contrast, the effects of neighborhood density in various psycholinguistic tasks are emergent properties of many models of speech production \citep[e.g.,][]{Luce1998,Luce2000} and perception \citep{Marslen-Wilson1984,Marslen-Wilson1987,McClelland1986,Norris1994,Norris2000}.

@@

This proposition has the form of an indicative conditional. While the antecedent of this condition, the presence or absence of intermediate ratings in a graded judgement task, is readily observable, it has no bearing on the truth or falsity of the conditional proposition. However, whereas a false consequent (e.g., an all-or-nothing concept) would potentially falsify the proposition, the consequent, a mental state, cannot be directly observed. If this is the case, then both the proposition and the consequent (the proposition that wordlikeness is itself gradient) fail the test of falsification, placing it beyond the scope of scientific inquiry.

@@

\citeauthor{Armstrong1983} note that if subjects produce intermediate ratings for these categories, then the naïve proposition is false, and this is what \citeauthor{Armstrong1983} find in their experiments. They ask subjects to rate, on a seven-point scale, the extent to which, e.g., certain odd counting numbers represent the concept ``odd number'', and find that speakers do make use of intermediate values, as shown in Figure \ref{agg}. In summary, the proposition that intermediate ratings indicate mental gradience is either untestable, or it is false, and is thus rejected.

@@

\citeauthor{Armstrong1983} do not view their results as evidence that subjects have an internal model of odd numbers which is at odds with the formal, all-or-nothing definition, and in fact, several additional experiments show that subjects deploy the extension of the formal definition in other tasks. Reviewing this study, \citet[215]{Schutze2011} writes that the experiments show that judgements are ``sensitive to factors other than our underlying competence''. In the case of wordlikeness, a case can be made that one such factor is similarity to existing words (e.g., \citealt{Bailey2001}, \citealt[][151, fn. 27]{LSLT}, \citealt{Greenberg1964}, \citealt{Ohala1986b}, \citealt{Schutze2005}, \citealt{Sendlmeier1987}, \citealt{Vitz1973}). However, the mere fact that subjects use intermediate ratings does not show that they do so with any systematicity, or that, e.g., the contrast between \emph{447} and \emph{7} with respect could be given any satisfying explanation. \citeauthor{Schutze2011} further suggests that gradient responses are the product of the task itself:

\begin{quote}
Putting it another way, when asked for gradient responses, participants will find some way to oblige the experimenter; if doing so is incompatible with the experimenter's actual question, they apparently infer that she must have really intended to ask something slightly different. \citep[215]{Schutze2011}
\end{quote}

This introduces a troubling possibility, that subjects ``oblige'' experimenters by introducing random noise to their categorical judgements when presented with a Likert task. Were this the case, the observed intermediate judgements are unlikely to yield information about speakers' knowledge of possible and impossible words. The alternative considered here is that there exists some model in which the gradience in ratings is systematic.

@@

\noindent There are now many computational models of wordlikeness which proport to do just this. An unfortunate defect of prior evaluations of these models, however, is that they adopt the naïve gradience hypothesis. \citet[382]{Hayes2008a} state that consider the ability to model gradient intuitions to be so important that they do not include any categorical models in their evaluation, as do many other authors. As a result, the literature contains no serious attempt to evaluate the gradience hypothesis.

@@

\citet{Greenberg1964} investigated wordlikeness using the technique of free magnitude estimation, a mechanism which has become increasingly popular among syntacticians \citep[e.g.,][]{Bard1996}. At the beginning of the experiment, the subject heard a recording of the word \emph{stick}. In subsequent trials, the subjects heard a nonce word and were asked to report ``how far would you say that is from English?'', with \emph{stick} at ``1''; subjects are told that a word that is ``twice as far from English'' as \emph{stick} should be scored ``2''. The data used here are from \citeauthor{Greenberg1964}'s Experiment B, in which 17 undergraduates were presented 17 stimuli in all. In addition to \emph{stick}, the stimuli include three other English words; these four items were excluded from further analyses, leaving 13 stimuli. As is standard practice in psychophysics \citep[e.g.,][]{Butler1987}, magnitudes were log-transformed prior to analysis.

@@

The binary baseline used here is a crude implementation of the null hypothesis that there are no gradient effects in wordlikeness judgements. To create such a baseline, it is necessary to distinguish between nonce words which contain a gross phonotactic violation and those which do not. As all stimuli here are monosyllables, this task can be further simplified by separately considering the two major subcomponents of the syllable, the onset and rime. Speakers are particularly adept as separating onsets from rimes \citep{Treiman1986,Treiman1995,Fowler1993}, and a large portion of phonotactic generalizations and violations can be localized to either onset or rime \citep[e.g.,][]{Borowsky1989,Fudge1969,Treiman1995,Kessler1997,Treiman2000}.

The baseline considers a nonce syllable to be well-formed if it consists of both an attested onset and an attested rime; the free combination of these two components is the only mechanism by which this model can generalize beyond attested words. It is surely the case that this is an ultimately insufficient model: \citet{Albright2009a} observes, for instance, that [ɛsp] is a well-formed rime, even though it is found in no word of English. This problem is more acute in languages with more permissive syllable structures than those of English. For instance, \citet{Fischer-Jorgensen1952} and \citet{Vogt1954} assert that there are many accidental gaps (i.e., possible but unattested structures) in the inventory of consonant clusters in Georgian, a language which admits as many as five adjacent consonants.\footnote{
Chapter \ref{clusters} 
%\citet{Gorman2012c}
reports a similar result for English syllable contact clusters.} Despite these defects, the baseline outperforms the gradient models in most contexts.

\begin{figure} \centering
\includegraphics{density.pdf}
\caption{Average ratings of individual nonce words, linearly transformed to fit the interval [0, 1], tend to clump into two groups with little overlap; words which consist of  attested onsets and rimes receive ratings near ceiling, whereas ratings of phonotactically invalid words are spread across the lower half of the spectrum.}
\label{dsn}
\end{figure}

@@

@@

\citeauthor{Hayes2008a} (\citeyear{Hayes2008a}; henceforth H\&W) develop a sophisticated model of phonotactic grammaticality which estimates a probability distribution over phoneme sequences by weighing constraints according to the principle of maximum entropy, following \citet{Goldwater2003} and \citet{Jager2007}. H\&W report that the predictions of their model are closely correlated with the \citet{Scholes1966} wordlikeness ratings. A direct replication of their predictions was attempted by using the software, model parameters, and training data as described in that study. Since the training of the maximum entropy model is inherently stochastic, producing slightly different outcomes on each run, the lowest scoring of ten runs is reported (H\&W:396), though in general there is not a great deal of variation between individual runs. One limitation of this model is that it is not feasible to score whole words, as the number of constraints which must be inspected grows exponentially as the span of possible constraints increases. Following H\&W and of \citet{Albright2009a}, who also applies the maximum entropy model to the \citet{Albright2003b} norms, the model is trained and scored only on stimulus onsets. However, as a consequence, the maximum entropy model performs particularly poorly on this data set, as many stimuli contain phonotactic violations in rime positions.

@@

\noindent \citet{Albright2009a} employs bigram probability to model wordlikeness judgements. While the focus of \citeauthor{Albright2009a}'s study is on developing a model which uses bigrams over phonological features rather than segments themselves, \citeauthor{Albright2009a}'s evaluation, which includes both the \citeauthor{Scholes1966} and \citeauthor{Albright2003b} data sets, finds an advantage for segmental bigrams. \citeauthor{Albright2009a} does not provide an implementation of the featural bigram model, nor does his study describe it in sufficient detail to allow for a new implementation, but segmental bigram scores for the \citeauthor{Albright2003a} data are reported in the appendix. As reported by \citeauthor{Albright2009a}, segmental bigrams outperform featural bigrams (see Table \ref{albrightimproved}).

\citeauthor{Albright2009a} estimates bigram probabilities using the method of maximum likelihood over types in the lexicon. The variant of segmental bigrams used here computes probabilities with a simple type of smoothing in which the count of all possible bigrams (including those never observed) are incremented by one. This technique is known as Laplace, or ``add one'' smoothing. This has the desirable effect that no nonce word is ever assigned a zero probability, and produces a small increase in the correlation between the \citeauthor{Albright2003b} wordlikeness norms compared with the maximum likelihood estimate (Table \ref{albrightimproved}). For all three data sets, this model also consistenly outperforms positional probability models defined by \citet{Vitevitch2004} and \citet{Vaden2009}. 
Given this and the fact that these model scores are highly correlatd with bigram probability \citep[][54]{Vitevitch1997}, they are not considered further. The bigram model consistently performs well in all the evaluations, and has the highest Spearman correlation with the \citeauthor{Greenberg1964} and \citeauthor{Scholes1966} data, and is frequently second place model to the binary baseline elsewhere.
